### 2024-03-28 doc/sections/mav-cut-elimination.tex
[ tex ] finish off the semantics and cut elimination sections

>>>>

Another consequence of Proposition \ref{prop:embedding-sem} is that
MAV is complete for the MAV-frame semantics:

\begin{theorem}\label{thm:completeness}
  MAV is complete for the MAV-frame semantics: if, for all MAV-frames
  $F$, $\sem{I} \sqsubseteq \sem{P}$ in
  $(\Chu(\ClosedLowerSet{F}, \ClosedDay{I}), \sqsubseteq)$, then
  $P \longrightarrow^* I$.
\end{theorem}

### 2024-03-28 doc/sections/mav-semantics.tex
[ tex ] more fixes
[ tex ] finish off the semantics and cut elimination sections

\bob{Short introduction to the section here}

\begin{itemize}
\item We construct algebraic models of MAV
\item MAV-algebras are algebras that MAV is sound for. In some
  sense, this is straightforward, but we identify precisely the
  structure that is needed.
\item The key structure is duoidality.
\item We develop some model theory here, and show that MAV-algebras
  can be induced by the weaker structure of MAV-frames. This is
  essential for the cut-free elimination proof.
\end{itemize}

@@

\begin{definition}
  A \emph{partial order monoid (pomonoid)} $(\bullet, i)$ on a poset
  $(A, \leq)$ comprises a binary operator $\bullet : A \times A \to A$
  that is monotone in both arguments and an element $i \in A$ such
  that the usual monoid laws hold. A \emph{commutative pomonoid} is a
  pomonoid where additionally $x \bullet y = y \bullet x$.
\end{definition}

\begin{definition}
  A commutative pomonoid $(\bullet, i)$ on a poset $(A, \leq)$ is
  \emph{residuated} if there is a function
  $\rightblackspoon : A \times A \to A$ such that $x \bullet y \leq z$
  iff $x \leq y \rightblackspoon y$.
\end{definition}

@@

\begin{definition}
  A \emph{$*$-autonomous partial order} is a structure
  $(A, \leq, \otimes, I, \lnot)$ where $(\otimes, I)$ is a pomonoid on
  $(A, \leq)$ and $\lnot : A^\op \to A$ is an anti-monotone and
  involutive operator on $A$, together satisfying
  $x \otimes y \leq \lnot z$ iff $x \leq \lnot (y \otimes z)$.  A
  $*$-autonomous partial order satisfies \emph{mix} if $\lnot I = I$.
\end{definition}

@@

\begin{definition}
  A pomonoid $(\bullet, i)$ is \emph{duoidal over} another pomonoid
  $(\lhd, j)$ on a partial order $(A, \leq)$ if the following
  inequalities hold:
  \begin{enumerate}
    \item $(w \lhd x) \bullet (y \lhd z) \leq (w \bullet y) \lhd (x \bullet z)$
    \item $j \bullet j \leq j$
    \item $i \leq i \lhd i$
    \item $i \leq j$
  \end{enumerate}
\end{definition}

@@

\begin{proposition}\label{prop:mav-algebra-consequences}
  Let $(A, \leq, \otimes, \lhd, I, \lnot)$ be a MAV-algebra.
  \begin{enumerate}
    \item There is another pomonoid structure $(\parr, I)$ on
          $(A, \leq)$, defined as
          $x \parr y = \lnot(\lnot x \otimes \lnot y)$.
    \item $(\otimes, I)$ and $(\parr, I)$ are linearly distributive
          \cite{cockett}:
          $x \otimes (y \parr z) \leq (x \otimes y) \parr z$.
    \item $(A, \leq)$ has binary joins, given by
          $x \oplus y = \lnot (\lnot x \with \lnot y)$.
    \item $\oplus$ distributes over $\otimes$:
          $x \otimes (y \oplus z) = ((x \otimes y) \oplus (x \otimes z)$.
    \item $\with$ distributes over $\parr$:
          $(x \parr z) \with (y \parr z) = (x \with y) \parr z$.
    \item $\lhd$ is duoidal over $\parr$:
          $(w \parr x)\lhd (y \parr z) \leq (w\lhd y) \parr (x\lhd z)$.
    \item $\lhd$ is duoidal over $\with$:
          $(w \with x)\lhd (y \with z) \leq (w \lhd y) \with (x\lhd z)$.
    \item $\oplus$ is duoidal over $\lhd$:
          $(w \lhd x) \oplus (y \lhd z) \leq (x \oplus y) \lhd (x \oplus z)$.
  \end{enumerate}
\end{proposition}

@@

\begin{definition}
  An \emph{MAV-frame} is a structure $(F, \leq, \parr, \lhd, I, +)$
  where $(F, \leq)$ is a partial order, $(F, \leq, \parr, I)$ is a
  commutative pomonoid, $(F, \leq, \lhd, I)$ is a pomonoid, $+$ is a
  binary monotone function on $(F, \leq)$, and these data satisfy the
  following inequalities:
  \begin{enumerate}
    \item $(w \lhd x) \parr (y \lhd z) \leq (w \parr y) \lhd (x \parr z)$
    \item $(x + y) \parr z \leq (x \parr z) + (y \parr z)$
    \item $(w \lhd x) + (y \lhd z) \leq (w + y) \lhd (x + z)$
    \item $I + I \leq I$
  \end{enumerate}
\end{definition}

@@

\begin{theorem}\label{thm:algebra-from-frame}
  If $(F, \leq, \parr, \lhd, I, +)$ is an MAV-frame, then
  $(\Chu(\ClosedLowerSet{F}, I), \sqsubseteq)$ has the structure of an
  MAV-algebra.
\end{theorem}

\begin{proof}
  Combining the propositions above, in the special case of the units
  of the two pomonoids being equal.
\end{proof}

>>>>

To prove the normalisation property for all MAV proofs, we use a
semantic technique that is akin to Okada's phase space method and to
Normalisation by Evaluation (NbE). We construct a semantics of the
whole proof system from the system of normal proofs. This semantics is
constructed in such a way that after interpreting a proof, the
(existence of) a normal form can be extracted (or \emph{read back} or
\emph{reified} in NbE terminology) from the semantic proof.

To our knowledge, the semantics of MAV have not been previously
studied in the literature. The rules are an extension of MALL, so an
\emph{MAV-algebra} will partly xbe a $*$-autonomous partial order
(Definition \ref{defn:star-autonomous}) with meets and joins. The
additional structure for the $\vP \vSeq \vQ$ connective satisfies the
conditions of a duoidal category \cite{hopf-book} (Definition
\ref{defn:duoidal}). We show that MAV is sound for MAV-algebras in
Theorem \ref{thm:soundness}.

To build MAV-algebras from normal proofs, we define the weaker notion
of an \emph{MAV-frame} (Definition \ref{defn:mav-frame}. We then show
that a combination of certain closed lower sets (Section
\ref{sec:closed-lower-sets}) and the Chu construction (Section
\ref{sec:chu}) construct an MAV-algebra from any MAV-frame. Much of
these constructions are well-known, but we have new results on lifting
the Day construction of monoids on closed lower sets and the
preservation of duoidal relationships that are required for MAV.

@@

\begin{definition}\label{defn:pomonoid}
  A \emph{partial order monoid (pomonoid)} $(\bullet, i)$ on a poset
  $(A, \leq)$ comprises a binary operator $\bullet : A \times A \to A$
  that is monotone in both arguments and an element $i \in A$ such
  that the usual monoid laws hold. A \emph{commutative pomonoid} is a
  pomonoid where additionally $x \bullet y = y \bullet x$.
\end{definition}

\begin{definition}\label{defn:residual}
  A commutative pomonoid $(\bullet, i)$ on a poset $(A, \leq)$ is
  \emph{residuated} if there is a function
  $\rightblackspoon : A \times A \to A$ such that $x \bullet y \leq z$
  iff $x \leq y \rightblackspoon y$.
\end{definition}

@@

\begin{definition}\label{defn:star-autonomous}
  A \emph{$*$-autonomous partial order} is a structure
  $(A, \leq, \otimes, I, \lnot)$ where $(\otimes, I)$ is a pomonoid on
  $(A, \leq)$ and $\lnot : A^\op \to A$ is an anti-monotone and
  involutive operator on $A$, together satisfying
  $x \otimes y \leq \lnot z$ iff $x \leq \lnot (y \otimes z)$.  A
  $*$-autonomous partial order satisfies \emph{mix} if $\lnot I = I$.
\end{definition}

@@

\begin{definition}\label{defn:duoidal}
  A pomonoid $(\bullet, i)$ is \emph{duoidal over} another pomonoid
  $(\lhd, j)$ on a partial order $(A, \leq)$ if the following
  inequalities hold:
  \begin{enumerate}
    \item $(w \lhd x) \bullet (y \lhd z) \leq (w \bullet y) \lhd (x \bullet z)$
    \item $j \bullet j \leq j$
    \item $i \leq i \lhd i$
    \item $i \leq j$
  \end{enumerate}
\end{definition}

@@

\begin{proposition}\label{prop:mav-algebra-consequences}
  Let $(A, \leq, \otimes, \lhd, I, \lnot)$ be a MAV-algebra.
  \begin{enumerate}
    \item There is another commutative pomonoid structure $(\parr, I)$ on
          $(A, \leq)$, defined as
          $x \parr y = \lnot(\lnot x \otimes \lnot y)$.
    \item $(\otimes, I)$ and $(\parr, I)$ are linearly distributive
          \cite{cockett}:
          $x \otimes (y \parr z) \leq (x \otimes y) \parr z$.
    \item $(A, \leq)$ has binary joins, given by
          $x \oplus y = \lnot (\lnot x \with \lnot y)$.
    \item $\oplus$ distributes over $\otimes$:
          $x \otimes (y \oplus z) = (x \otimes y) \oplus (x \otimes z)$.
    \item $\with$ distributes over $\parr$:
          $(x \parr z) \with (y \parr z) = (x \with y) \parr z$.
    \item $\lhd$ is duoidal over $\parr$:
          $(w \parr x)\lhd (y \parr z) \leq (w\lhd y) \parr (x\lhd z)$.
    \item $\lhd$ is duoidal over $\with$:
          $(w \with x)\lhd (y \with z) \leq (w \lhd y) \with (x\lhd z)$.
    \item $\oplus$ is duoidal over $\lhd$:
          $(w \lhd x) \oplus (y \lhd z) \leq (x \oplus y) \lhd (x \oplus z)$.
  \end{enumerate}
\end{proposition}

@@

\begin{definition}\label{defn:mav-frame}
  An \emph{MAV-frame} is a structure $(F, \leq, \parr, \lhd, I, +)$
  where $(F, \leq)$ is a partial order, $(F, \leq, \parr, I)$ is a
  commutative pomonoid, $(F, \leq, \lhd, I)$ is a pomonoid, $+$ is a
  binary monotone function on $(F, \leq)$, and these data satisfy the
  following inequalities:
  \begin{enumerate}
    \item $(w \lhd x) \parr (y \lhd z) \leq (w \parr y) \lhd (x \parr z)$
    \item $(x + y) \parr z \leq (x \parr z) + (y \parr z)$
    \item $(w \lhd x) + (y \lhd z) \leq (w + y) \lhd (x + z)$
    \item $I + I \leq I$
  \end{enumerate}
\end{definition}

@@

The propositions in the preceeding three sections together prove that
every MAV-frame yields an MAV-algebra:

\begin{theorem}\label{thm:algebra-from-frame}
  If $(F, \leq, \parr, \lhd, I, +)$ is an MAV-frame, then
  $(\Chu(\ClosedLowerSet{F}, \ClosedDay{I}), \sqsubseteq)$ has the structure of an
  MAV-algebra.
\end{theorem}

With this theorem we can define a notion of validity in MAV in terms
of truth in all MAV-frame generated algebras. By Theorem
\ref{thm:soundness}, MAV is sound for this notion of validity:

\begin{theorem}\label{thm:mavframe-soundness}
  MAV is sound for the MAV-frame semantics: if $P \longrightarrow^* I$
  then for all MAV-frames $F$, $\sem{I} \sqsubseteq \sem{P}$ in
  $(\Chu(\ClosedLowerSet{F}, \ClosedDay{I}), \sqsubseteq)$.
\end{theorem}
